---
title: "inferential_assignment_1_part2"
author: "treepruner"
date: "September 24, 2015"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Description
This part of the exercise analyzes the effect of citamin C on tooth growth in guinea pigs. The  ToothGrowth data in the R datasets package. 

The response is the length of odontoblasts (teeth) in each of 10 guinea pigs at each of three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods (orange juice or ascorbic acid).

## Assumptions
I assumed that the guinea pigs were randomly assigned to the treatment groups and each guinea pig only received on type of supplement, e.g., they were NOT paired observations.


## Exploratory Data Analyses
Determine the variables, class and sample values.
```{r echo = FALSE, message = FALSE} 
library(ggplot2)
library(dplyr)
data("ToothGrowth")
str(ToothGrowth)
```


```{r echo = FALSE, message=FALSE, warning=FALSE,  fig.height = 3.5, fig.width = 3.5, }
# Convert the variable dose from a numeric to a factor variable
ToothGrowth$dose <- as.factor(ToothGrowth$dose)

g <- ggplot(ToothGrowth, aes(x = len, color = dose))
g <- g + geom_density()
g <- g + facet_grid(. ~ supp)
g <- g + ggtitle( "Examining Shape of Data")
g <- g + xlab ("Tooth Length") + ylab("Density") 
g 


oj_vj2 <- ggplot(ToothGrowth, aes(x = factor(dose), y = len, color = supp )) + geom_boxplot()
oj_vj2 <- oj_vj2 + scale_y_continuous(breaks=seq(0,35,2))
oj_vj2 <- oj_vj2 + facet_grid(.~supp)
oj_vj2 <- oj_vj2 + ggtitle( "Boxplot by Dose and Supplement")
oj_vj2 <- oj_vj2 + xlab ("Dose") + ylab("Tooth Length") 
oj_vj2 


```

```{r echo = FALSE, message=FALSE, warning=FALSE,  fig.height = 2.5, fig.width = 3.5, }

oj_vj1 <- ggplot(ToothGrowth, aes(x = dose, y = len, color =  supp)) + geom_point( alpha = .2, size = 4)
oj_vj1 <- oj_vj1 + facet_grid(.~supp)
oj_vj1 <- oj_vj1 + xlab ("Dose") + ylab("Tooth Length") 
oj_vj1 <- oj_vj1 + ggtitle("Comparison by Supplement Type")
oj_vj1

dose2 <- ToothGrowth %>% filter(dose ==2)
g3 <- ggplot(dose2, aes(x = len, color = supp ))
g3 <- g3 + geom_density()
g3 <- g3 + facet_grid(. ~ supp)
g3 <- g3 + labs(x = "Supplement" )     
g3 <- g3 + labs(y = "Tooth Growth Length") 
g3 <- g3 + ggtitle( "Tooth Growth at Highest Dose")
g3
```


## Data Summary
```{r echo = FALSE, message = FALSE} 
Treatment.Summary  <- ToothGrowth %>%  group_by(supp, dose) %>% summarise(n(), mean(len), median(len), IQR(len), sd(len), min(len), max(len))
Treatment.Summary 
```

## Use Confidence Intervals and/or Hypothesis on ToothGrowth

### Test 1
The highest dose of each supplement has the highest growth. At  this level the mean and the median are approximately the same  for each supplement, but VC has a wider spread than OJ. Is this difference significant? Perform a T Test on the dose2 subset.

```{r echo = FALSE, message = FALSE} 
t.test(len ~ supp, paired = FALSE, var.equal = FALSE, data = dose2)
```
The 95% confidence interval for Test 1 contains 0, so we can not reject the NULL hypothesis that there is no different between the means of the groups.

### Test 2
There is an overlap in the boxplots for OJ at doses 1 and 2. Is this difference significant?
```{r echo = FALSE, message = FALSE} 
OJ_1_2 <- ToothGrowth %>% filter(dose != .5 & supp == "OJ")
t.test(len ~ dose, paired = FALSE, var.equal = FALSE, data = OJ_1_2)
```
The 95% confidence interval for Test 2 does not contain 0 and the p value is less than .05, so these results are significantly  different from each other.


### Test 3
There is an overlap in the boxplots dose .5. Is this difference significant?
```{r echo = FALSE, message = FALSE} 
low <- ToothGrowth %>% filter(dose == .5 )
t.test(len ~ supp, paired = FALSE, var.equal = FALSE, data = low)
```

The 95% confidence interval for Test 3 does not contain 0 ad the p value is less than .05, so these resuls are significantly different from each other.

## Conclusions
There sample sizes were very small, only 10 observations per supplement and dose combinations. The T-test is valid for small sample sizes.

If you want to maximize guinea pig tooth growth either orange juice or Vitamin C at dose = 2mg are effective and there isn't a significant difference in their performance. Choose the supplement based upon price or availability.

If you want to minimize guinea pig tooth growth, Vitamin C at the 0.5 mg dose is significantly lower than orange juice at 0.5 mg. 
 


